Embracing the Laws of Physics: \\ Reversible Models of Computation

### 2017-12-08 survey/survey.tex
add lots of comments to Data II, III and IV.

Pi level 1

@@

name the isomorphisms

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data III: Reversible Programs between Structured Finite Types}

Pi level 2

@@

>>>>

\jc{wavefront}

Introduce the usual types. Curry-Howard, as usual. Problem with it, i.e.
it is merely about inhabitation, which is NOT reversible. Show examples
with A+A and A*A. Our deformations need to be reversible, so they need
to induce type isomorphisms. So while our types are as usual, the
first level of transformations on data will change, and be isomorphisms.
The type-erasure will, naturally, be permutations. But where do the
isomorphisms come from?  Well, inhabitants of types with (0,1,+,*)
form a semiring. So, still inspired by Curry-Howard, look to the
proof-terms of semiring identities -- and voila, a nice set of
isomorphisms! We can name these isomorphisms, AND give them
implementations. Gives us a nice, reversible PL for
structured finite types. We can it Pi.

We actually have several choices. We can make Pi structurally reversible
(i.e. there is a syntactic transformation on Pi which maps combinators
to combinators which witnesses reversibility), or build in a `reverse'
combinator which flips things around. The latter lets us make the
language much smaller, but then \emph{programs} in that language are longer.
[Bias: since there will be many more programs written, the fact that the
language definition, and its processors, is larger, is a one-time cost.
Not that big a deal. So pick the version whose meta-theoretical
properties are obvious. It is thus worthwhile to have reversibility be
structural.]

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data III: Reversible Programs between Reversible Programs}

Pi level 2.

Take categorification seriously. Types are indeed groupoids, and
endo-isomorphisms inhabit the Identity 'type' of a type. So + and *
must be operations on groupoids. Look at the category of (small)
categories for inspiration. So we have two monoidal structures.
These are not co-product and product in Groupoid. But this is actually
a good thing.

So we end up looking, semantically, at weak Rig Groupoid. Syntactically,
we take the easiest way there: simply make every coherence isomorphism into
a combinator. Huge proof obligation that this induces actual transformations
on Pi programs. Luckily, it does.  But not straightforwardly -- ``incorrect''
definitions of some Pi combinators (such as those for 0*x = 0) can lead to
some level 2 morphisms not working.

The interpretation on permutations is unclear, because the Pi representation
has so much more structure, and permutations too have many different 
representations. So it is unclear what maps to what.

@@

DANGER: here, we leave the safe land of ``preservation of information'',
because partiality is not reversible. So while trace gives a reversible
program as output, it itself isn't a reversible operation. Like the
demon's brain in Maxwell's thought experiment, trace irreversible work!
So trace, which is fundamental to all work with recursive types, is
only reversible "at level 1". So we need to replace trace with some other
operation for level 2 reversibility to also hold -- and this is an open
problem.
